A Gentle Introduction into Structural Causal Models

### 2021-06-10 paper_scm.Rmd
paper

---
title: "A Gentle Introduction into Structural Causal Models"
author: 
- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
- Department of Statistics, Ludwig Maximilian University Munich, Germany
date: "June 2021"
output: 
  pdf_document:
        number_sections: yes
        fig_caption: yes
link-citations: yes
linkcolor: blue
fontsize: 12
fontfamily: mathpazo 
linestretch: 1.5
citation_package: --biblatex
bibliography: [ref.bib]
csl: ieee.csl
nocite: '@*'
---

@@

Section 2 introduces the assumptions in causal modelling. Further, section 2 discusses different assumptions in causal modelling. Section 3
Section 4 addresses Pearls Causal Hierachy.
Section 5 provides a brief introduction into graphical models.
Section 6 focuses on the intersection of SCMs and the perception of time.

# SCM Example

The structural causal model consists of a set of equations.
These equations are asymmetric, because they are not bi-directional.
We cannot change the sides of the equation as one can do with regular equations [@pearl2012causal].

$S:= f_S(U_S)$
$E:= f_E(U_E)$
$C:= f_C(S,E,U_C)$

$\{S\}$ - Smoking
$\{C\}$ - Cancer
$\{E\}$ - Exercise

A structural causal model contains an underlying graphical model [@hardtrecht]. 
The most popular graphical model is the directed acyclic graph or in short DAG.
A DAG entails nodes and edges. 
Nodes represent our different variables. 
Edges depict the assignment equations. 
All edges are directed in the DAG.
The graph is called acyclic if no variable causes (directly or indirectly) itself [@morgan_winship_2014].
There is a hierachy within DAGs.

```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left of=A](UA){$U_A$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,left = 1cm of T](UT){$U_T$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (T);
\draw[->] (UA) -- (A);
\draw[->] (UT) -- (T);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

Probabilisitic Representation

>>>>

---
title: "A Gentle Introduction into Structural Causal Models"
author: 
- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
- Department of Statistics, Ludwig Maximilian University Munich, Germany
date: "June 2021"
output: 
  pdf_document:
        number_sections: yes
        fig_caption: yes
link-citations: yes
linkcolor: blue
fontsize: 12
fontfamily: mathpazo 
linestretch: 1.5
citation_package: --biblatex
bibliography: [ref.bib]
nocite: '@*'
---

@@

Section 2 introduces the assumptions in causal modelling
Section 3 addresses Pearls Causal Hierachy.
Section 4 provides a brief introduction into graphical models.
Section 5 focuses on the intersection of SCMs and the perception of time.

# SCM Foundation

The structural causal model consists of a set of equations.
These equations are asymmetric assignments, because they are not bi-directional.
We cannot change the sides of the equation as one can do with regular equations [@pearl2012causal].

We can set up a generic example with tree variables. 
One common example from epidemiology is looking at the impact of problem behaviour and health enhancing behaviour.
Problem behaviour is a latent variable, but we can use observational data to characterize problem behaviour.
One example of problem behaviour is smoking 
To estimate smoking, we can for instance ask participants how frequently they smoke and then describe p.

As we can see, these functions are driven by underling latent variables.
These latent factors are the foundation of the structural causal model [@hardtrecht; @pearl2009causality; @pearl2012causal].

\begin{equation}
S:= f_S(U_S)
\end{equation}
\begin{equation}
E:= f_E(U_E)
\end{equation}
\begin{equation}
C:= f_C(S,E,U_C)
\end{equation}

where: 
$\{S\}$ - Frequency of Smoking
$\{E\}$ - Frequency of Exercise
$\{C\}$ - Lung Cancer

Every structural causal model contains an underlying graphical model [@hardtrecht]. 
The most popular graphical model is the directed acyclic graph or in short DAG.
A DAG entails nodes and edges. 
Nodes represent our different variables. 
Edges depict the assignment equations. 
All edges are directed in the DAG.
The graph is called acyclic if no variable causes (directly or indirectly) itself [@morgan_winship_2014].
There is a hierachy within DAGs.

```{tikz,fig.cap="Structural Causal Model" ,fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (S){S};
\node[latent,left of=S](US){$U_S$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,left = 1cm of T](UT){$U_T$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (T);
\draw[->] (UA) -- (A);
\draw[->] (UT) -- (T);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

For reference, we can also display this model based on other models. 
One example is the causal bayesian network, which uses conditional probabilities instead of functions to describe the relationship between variables. 
In the probabilistic representation, we ignore latent factors.
